Neuroevolution in a simple two-player game situation

### 2019-01-07 documentation/paper.tex
More small fixes
Title
Last fixes

\date{\today}
\title{Really Good Title}
\subtitle{Genetic Algorithms and Evolutionary Computing Project}
\author{Felix Bartel and D\'avid Kerekes}

@@

\begin{figure}[H]
\begin{tabular}{c|ccccccc}
vs. &
\texttt{Axji-0-2} &
\texttt{COM\_11} &
\texttt{GintonicV9} &
\texttt{Hypo14} &
\texttt{reduced} &
\texttt{Union} &
\texttt{trainer} \\
\hline
\texttt{Axji-0-2} & - & 0 & 0 & 4 & 0 & 0 & 10 \\
\texttt{COM\_11} & 21 & - & 9 & 19 & 6 & 16 & 20 \\
\texttt{GintonicV9} & 21 & 12 & - & 19 & 6 & 16 & 20 \\
\texttt{Hypo14} & 17 & 2 & 2 & - & 2 & 5 & 15 \\
\texttt{reduced} & 21 & 19 & 15 & 19 & - & 19 & 21 \\
\texttt{Union} & 21 & 14 & 5 & 16 & 2 & - & 20 \\
\texttt{trainer} & 11 & 0 & 1 & 6 & 0 & 1 & -
\end{tabular}
\label{tab:bot_vs_bot}
\caption{Scores of bots competing against each other}
\end{figure}

@@

For the whole experiment, we fixed the size of the neural network to $6$ input nodes, one hidden layer with $7$ nodes and $2$ output nodes. As can be seen later, this proved sufficient for a surprisingly high performance by the network. The input nodes got fed with the x and y positions of the player, it's relative position compared to the ball, and the velocity vector of the ball.
The output nodes consisted of two neurons with tanh activations, one responsible for going left or right in the next timestep (if the output was below or above of threshold values $t_l = 0.49, t_r = 0.51$) and the other for jumping (above threshold $t_j = 0.7$).

@@

\todo{overall: not very pretty graphs <<-- Why do we decided to do self adaptation?}

We first implemented a method similar to (1) and (4) from \cite{self_adapt}, giving each individual a mutation rate and a sigma value. These were mutated when the individual was selected for mutation. Their initialization was also random, in the range of [0.25, 0.75] for the mutation rate and [0, 0.03] for the sigma values.

@@

\todo{too much diversity <<-- what does this mean?}

In this scenario \ref{fig:self_adapt_1_self_training_2_oldrun} can be seen battling for being the better species in the beginning, but then this seems to converge. A quick manual look at the games confirms however, that they are both still actively learning. When left alone to train for 8000 generations (figure \ref{fig:self_adapt1_self_training_2}) we get the best results so far (figure \ref{fig:self_training_2}). The only bot capable of matching this network is the strongest \texttt{reduced} bot, but it never trained against any of the built-in bots directly. (In fact, this network was too strong even for us humans to beat in the game.)

@@

In the proposal, we listed more possible mutation and crossover operators from \cite{montana1989training}, that we later decided not to implement. The main reason behind is that these extra operators were based on modifying network architecture or size. After the first promising test runs, we decided on a fixed size, fully-connected architecture supported by hyperparameter self-adaptation, so we would have the compute time to pursue the more interesting self-training approach instead of searching the hyperparameter space of multiple architecture operators. Any other connection scheme, or smaller network can be represented by setting the correct weights to zero in our bigger network.

\todo{everything we have not done}
\todo{train the self trained bot with reduced so we maybe get a perfect score of 12}

We would like to thank \todo{xyz lab or something} TU-Chemnitz for letting us use their compute servers for our experiments.

>>>>

\date{\today}
\title{Neuroevolution in a simple two-player game situation}
\subtitle{Genetic Algorithms and Evolutionary Computing Project}
\author{Felix Bartel and D\'avid Kerekes}

@@

\begin{figure}[H]
\begin{tabular}{c|ccccccc}
vs. &
\texttt{Axji-0-2} &
\texttt{COM\_11} &
\texttt{GintonicV9} &
\texttt{Hypo14} &
\texttt{reduced} &
\texttt{Union} &
\texttt{trainer} \\
\hline
\texttt{Axji-0-2} & - & 0 & 0 & 4 & 0 & 0 & 10 \\
\texttt{COM\_11} & 21 & - & 9 & 19 & 6 & 16 & 20 \\
\texttt{GintonicV9} & 21 & 12 & - & 19 & 6 & 16 & 20 \\
\texttt{Hypo14} & 17 & 2 & 2 & - & 2 & 5 & 15 \\
\texttt{reduced} & 21 & 19 & 15 & 19 & - & 19 & 21 \\
\texttt{Union} & 21 & 14 & 5 & 16 & 2 & - & 20 \\
\texttt{trainer} & 11 & 0 & 1 & 6 & 0 & 1 & -
\end{tabular}
\label{tab:bot_vs_bot}
\caption{Wins of bots competing against each other}
\end{figure}

@@

For the whole experiment, we fixed the size of the neural network to $6$ input nodes, one hidden layer with $7$ nodes and $2$ output nodes. As can be seen later, this proved sufficient for a surprisingly high performance by the network. The input nodes got fed with the x and y positions of the player, it's relative position compared to the ball, and the velocity vector of the ball.
The output nodes consisted of two neurons with $\tanh$ activations, one responsible for going left or right in the next timestep (if the output was below or above of threshold values $t_l = 0.49, t_r = 0.51$) and the other for jumping (above threshold $t_j = 0.7$).

@@

Since running the algorithm is still quite resource intensive, we wanted to focus on improving the algorithm itself instead of hyperparameter optimization.

We first implemented a self adaptation method similar to (1) and (4) from \cite{self_adapt}, giving each individual a mutation rate and a sigma value. These were mutated when the individual was selected for mutation. Their initialization was also random, in the range of [0.25, 0.75] for the mutation rate and [0, 0.03] for the sigma values.

@@

In this scenario (figure \ref{fig:self_adapt_1_self_training_2_oldrun}) the two species can be seen battling for higher fitness in the beginning, but then this seems to converge. A quick manual look at the games confirms however, that they are both still actively learning. When left alone to train for 8000 generations (figure \ref{fig:self_adapt1_self_training_2}) we get the best results so far (figure \ref{fig:self_training_2}). The only bot capable of matching this network is the strongest \texttt{reduced} bot, but it never trained against any of the built-in bots directly. (In fact, our last neural network was too strong even for us humans to beat in the game.)

@@

A video of our two self-trained bots playing games can be found at \texttt{documentation/two_species.mp4}

In the proposal, we listed more possible mutation and crossover operators from \cite{montana1989training}, that we later decided not to implement. The main reason behind is that these extra operators were based on modifying network architecture or size. After the first promising test runs, we decided on a fixed size, fully-connected architecture supported by hyperparameter self-adaptation, so we would have the compute time to pursue the more interesting self-training approach instead of searching the hyperparameter space of multiple architecture operators. Any other connection scheme, or smaller network can be represented by setting the correct weights to zero in our bigger network. We also skipped implementing tournament-selection because of time constraints.

In the future we would like to improve on our best network by combining self train and the trainers, in order to produce an undefeated champion of the game. Since the project is open-source, we also plan on contributing some of our agents back to the original.

We would like to thank Faculty of Mathematics of TU-Chemnitz for letting us use their compute servers for our experiments.
